---
title: "Supplementary material"
brand: false
---

```{r}
#| include: false

library(knitr)
library(tidyverse)
library(brms)
library(tidybayes)
library(cmdstanr)
library(glue)
library(forcats)
library(stringr)
library(ggridges)
library(posterior)
library(tidytext)
library(gt)
library(ggnewscale)
library(stringr)
library(latex2exp)
library(ggforce)
library(readxl)
library(patchwork)
library(gganimate)
library(distributional)
library(reshape2)
library(scales)



sigfig <- function(vec, n=3){ 
  ### function to round values to N significant digits
  # input:   vec       vector of numeric
  #          n         integer is the required sigfig  
  # output:  outvec    vector of numeric rounded to N sigfig
  
  formatC(signif(vec,digits=n), digits=n,format="fg", flag="#") 
}

theme_set(theme_ggdist())
theme_update(text = element_text(size = 24),
             plot.subtitle=element_text(hjust=0.5))

# path <- cmdstanr::cmdstan_path()
# set_cmdstan_path(path = "C:/Users/Lynx/.cmdstan/cmdstan-2.36.0")
set_cmdstan_path(path = Sys.getenv("CMDSTAN_PATH"))

knitr::knit_engines$set(
  stan = function(options) {
    code <- knitr:::one_string(options$code)
    out <- options$output.var
    if (!is.character(out) || length(out) != 1L) stop(
      "the chunk option output.var must be a character string ",
      "providing a name for the returned `CmdStanModel` object."
    )
    if (options$eval) {
      if (options$cache) {
        cache_path <- knitr:::valid_path(
          options[["cache.path"]],
          options$label
        )
        if (!dir.exists(cache_path)) {
          dir.create(cache_path)
        }
        model_file <- tempfile(
          fileext = ".stan",
          tmpdir = cache_path
        )
        cat(code, "\n", file = model_file)
      } else {
        model_file <- write_stan_file(code)
      }
      csm <- cmdstan_model(model_file)
      assign(out, csm, envir = knit_global())
    }
    engine_output(options, code, '')
  }
)



```

```{r}
#| include: false

df0 <- read_xlsx("D://workR/Blog/Projects/AZV_HCC/SCZ/SCZ_paired_scores2.xlsx")

df <- df0  %>% 
  janitor::clean_names() %>% 
  mutate(across(c(panss:cgi_2, pss), as.numeric),
         visit = factor(visit, levels = c("V1", "V2", "FU"))) %>% 
  separate_wider_delim(ctq, names = paste0("ctq_", 1:6), delim = ",", too_few = "align_start") %>% 
  mutate(across(ctq_1:ctq_6, as.numeric))

hcc0 <- read_csv("D://workR/Blog/Projects/AZV_HCC/SCZ/SCZ_HCC_MP.csv")

hcc <- hcc0 %>% 
  select(-c(group, MP)) %>% 
  mutate(visit = toupper(visit))

visit_levels <- c("V1", "V2", "FU")

df_hcc <- df %>% 
  full_join(hcc, by = c("kod", "visit"), relationship = "many-to-many") %>% 
  mutate(visit = factor(visit, levels = c("V1", "V2", "FU"))) %>%
  mutate(date_total = if_else(is.na(segment) & visit == "V1", -7, date_total)) %>% 
  mutate(date_total = if_else(is.na(segment) & visit == "V2", -6, date_total)) %>% 
  mutate(date_total = if_else(is.na(segment) & visit == "FU", -1, date_total)) %>% 
  mutate(date_total = if_else(kod == "H-YP8YL", date_total + 0, date_total)) %>% 
  group_by(kod) %>%
  mutate(
    visit_spec = paste(
      visit_levels[
        visit_levels %in% visit[!is.na(segment)]
      ],
      collapse = ", "
    )
  ) %>% 
  mutate(date_total = case_when(visit_spec == "V1" ~ if_else(!is.na(segment), date_total - 7, -1 - 7),
                                visit_spec == "V1, V2" ~ if_else(!is.na(segment), date_total - 6, -1 - 6),
                                visit_spec == "V2" ~ if_else(!is.na(segment), date_total - 6, -1 - 6),
                                # visit_spec == "V1, FU" ~ if_else(!is.na(segment), date_total - 7, -1 - 7),
                                # visit_spec == "V2, FU" ~ if_else(!is.na(date_total), date_total - 6, -1 - 6),
                                # visit_spec == "FU" ~ if_else(!is.na(date_total), date_total - 1, -1 - 1),
                                .default = date_total)) %>% 
  group_by(visit) %>% 
  mutate(visit_mark = quantile(date_total, 0.94, na.rm = TRUE)) %>%
  ungroup() %>% 
  mutate(z_panss = scale(panss)[, 1],
         log_hcc = log(HCC),
         z_log_hcc = scale(log_hcc)[, 1],
         z_pss = scale(pss)[, 1]) %>% 
  group_by(kod, visit) %>% 
  mutate(z_panss = if_else(row_number() == 1, z_panss, NA_real_),
         z_pss = if_else(row_number() == 1, z_pss, NA_real_)) %>%
  identity()
  

```


#### Causal structure


```{mermaid}
%%| fig-width: 6.5
%%| fig-align: center

flowchart LR

    SCZ[SCZ]

    Z((Latent Process))
    CTQ[CTQ]

    HCC[HCC]
    PANSS[PANSS]
    PSS[PSS]

    SCZ --> |"<span style='font-size:30px; color:#ff0055;'>t<sub>0</sub></span>"| Z
    CTQ --> |"<span style='font-size:30px; color:#ff0055;'>λ</span>"| Z

    Z --> HCC
    Z --> PANSS
    Z --> PSS
    
    style SCZ fill:#f55,stroke:#333,stroke-width:2px
    style CTQ fill:#f55,stroke:#333,stroke-width:2px
    style Z fill:#bbf,stroke:#333,stroke-width:2px,stroke-dasharray: 5 5
    style HCC fill:#9f9,stroke:#333,stroke-width:2px
    style PANSS fill:#9f9,stroke:#333,stroke-width:2px
    style PSS fill:#9f9,stroke:#333,stroke-width:2px

```



#### Model Formulation

##### 1. Latent Childhood Trauma Facets
For each subject $i$, the joint distribution of their underlying exposure to abuse ($c_{i,1}$) and neglect ($c_{i,2}$) is drawn from a centered multivariate normal distribution capturing population variance and structural trait correlation:

$$
\begin{align*}
\begin{bmatrix} c_{i,1} \\ c_{i,2} \end{bmatrix} &\sim \text{Multivariate-Normal} \left( \begin{bmatrix} 0 \\ 0 \end{bmatrix}, \; \boldsymbol{\Sigma}_c \right) \\
\boldsymbol{\Sigma}_c &= \text{diag}(\boldsymbol{\sigma}_c) \, \boldsymbol{\Omega}_c \, \text{diag}(\boldsymbol{\sigma}_c)
\end{align*}
$$

Where $\boldsymbol{\sigma}_c = [\sigma_{\text{abuse}}, \sigma_{\text{neglect}}]^T$ maps the population scales, and $\boldsymbol{\Omega}_c$ represents the trait correlation matrix.

##### 2. CTQ Measurement (Emission) Layer
The individual's true latent trauma states directly manifest across the five observed subscales of the Childhood Trauma Questionnaire. The Abuse factor ($c_{i,1}$) drives emotional, physical, and sexual dimensions, while the Neglect factor ($c_{i,2}$) drives emotional and physical neglect indicators:

$$
\begin{align*}
y_{m}^{(\mathrm{EA})} &\sim \text{Normal}\left(\alpha_{\mathrm{EA}} + c_{\text{subj}[m], 1}, \; \sigma_{\mathrm{EA}}\right) \quad \text{[Emotional Abuse]} \\
y_{m}^{(\mathrm{PA})} &\sim \text{Normal}\left(\alpha_{\mathrm{PA}} + c_{\text{subj}[m], 1}, \; \sigma_{\mathrm{PA}}\right) \quad \text{[Physical Abuse]} \\
y_{m}^{(\mathrm{SA})} &\sim \text{Normal}\left(\alpha_{\mathrm{SA}} + c_{\text{subj}[m], 1}, \; \sigma_{\mathrm{SA}}\right) \quad \text{[Sexual Abuse]} \\
y_{m}^{(\mathrm{EN})} &\sim \text{Normal}\left(\alpha_{\mathrm{EN}} + c_{\text{subj}[m], 2}, \; \sigma_{\mathrm{EN}}\right) \quad \text{[Emotional Neglect]} \\
y_{m}^{(\mathrm{PN})} &\sim \text{Normal}\left(\alpha_{\mathrm{PN}} + c_{\text{subj}[m], 2}, \; \sigma_{\mathrm{PN}}\right) \quad \text{[Physical Neglect]}
\end{align*}
$$

##### 3. Latent Dynamics & Personalized Reversion
The latent dysregulation state $z_i(t)$ transitions over irregular time intervals ($\Delta t = t - t'$) according to a continuous-time [(Ornstein–Uhlenbeck)](https://en.wikipedia.org/wiki/Ornstein%E2%80%93Uhlenbeck_process) engine. The individual recovery speed $\lambda_i$ forces a strict positive bound via the structural baseline floor $\lambda_0$:


$$
\begin{align*}
z_i(t) \mid z_i(t') &\sim \text{Normal} \left(e^{-\lambda_i \Delta t} \, z_i(t'),\; \sigma_z \sqrt{1-e^{-2\lambda_i \Delta t}} \right) \\
\lambda_i &= \lambda_0 + \exp\left(\gamma_{\text{abuse}} \, c_{i,1} + \gamma_{\text{neglect}} \, c_{i,2}\right)
\end{align*}
$$

##### 4. Longitudinal Process States and Clinical Measurements
The state trajectory begins at a unique initial tracking baseline anchored to a matched-pair intercept ($u_{\text{pair}}$) and shifted by diagnostic status (SCZ):

$$
\begin{align*}
z_i(t_0) &\sim \text{Normal}\left(u_{\text{pair}[i]} + \beta_{\mathrm{SCZ}} \, \mathrm{SCZ}_i, \; 1\right) \\
u_j &\sim \text{Normal}(0, \sigma_{\mathrm{pair}})
\end{align*}
$$

The evolving state $z_i(t)$ projects concurrent signals down into both historical biological archives and acute clinical cross-sections:

$$
\begin{align*}
y_m^{(\mathrm{HCC})} &\sim \text{Student-t}\left(\nu = 6, \; a_{\mathrm{HCC}} + b_{\mathrm{HCC}}\, z_{\text{state}[m]}, \; \sigma_{\mathrm{HCC}}\right) \\
y_m^{(\mathrm{PANSS})} &\sim \text{Normal}\left(a_{\mathrm{PANSS}} + b_{\mathrm{PANSS}} \, z_{\text{state}[m]}, \; \sigma_{\mathrm{PANSS}}\right) \\
y_m^{(\mathrm{PSS})} &\sim \text{Normal}\left(a_{\mathrm{PSS}} + b_{\mathrm{PSS}} \, z_{\text{state}[m]}, \; \sigma_{\mathrm{PSS}}\right)
\end{align*}
$$

##### 5. Prior Distributions

$$
\begin{align*}
\lambda_0, \, \gamma_{\text{abuse}}, \, \gamma_{\text{neglect}}, \, \beta_{\mathrm{SCZ}} &\sim \text{Normal}(0, 1) \\
\alpha_{\mathrm{EA}}, \, \alpha_{\mathrm{PA}}, \, \alpha_{\mathrm{SA}}, \, \alpha_{\mathrm{EN}}, \, \alpha_{\mathrm{PN}} &\sim \text{Normal}(0, 1) \\
a_{\mathrm{PANSS}}, \, a_{\mathrm{PSS}}, \, a_{\mathrm{HCC}} &\sim \text{Normal}(0, 0.5) \\
b_{\mathrm{PANSS}}, \, b_{\mathrm{PSS}}, \, b_{\mathrm{HCC}} &\sim \text{Normal}(0, 1) \\
\sigma_z, \, \sigma_{\mathrm{pair}}, \, \sigma_{\mathrm{HCC}}, \, \sigma_{\mathrm{PANSS}}, \, \sigma_{\mathrm{PSS}} &\sim \text{Exponential}(1) \\
\sigma_{\mathrm{EA}}, \, \sigma_{\mathrm{PA}}, \, \sigma_{\mathrm{SA}}, \, \sigma_{\mathrm{EN}}, \, \sigma_{\mathrm{PN}} &\sim \text{Exponential}(1) \\
\boldsymbol{\sigma}_c &\sim \text{Exponential}(1) \\
\boldsymbol{\Omega}_c &\sim \text{LKJ}(\eta = 2)
\end{align*}
$$

#### Stan code

The mathematical model is rewritten in the Stan language.

```{stan scz, output.var="scz_model", cache=FALSE}
#| code-fold: false
#| echo: true

data{

int <lower=1> N_state;
array[N_state] int <lower=1> subj_id;
int <lower=1> N_pair;
array[N_state] int <lower=1> pair_id;
vector[N_state] time;
array[N_state] int <lower=0, upper=1> is_first;
array[N_state] int <lower=0, upper=1> scz;

int <lower=1> N_hcc;
array[N_hcc] int <lower=1> state_hcc;
vector[N_hcc] y_hcc;

int <lower=1> N_panss;
array[N_panss] int <lower=1> state_panss;
vector[N_panss] y_panss;

int <lower=1> N_pss;
array[N_pss] int <lower = 1> state_pss;
vector[N_pss] y_pss;

int <lower=1> N_ctq;
int <lower=1> N_subj;
array[N_ctq] int <lower=1> subj_ctq;

vector[N_ctq] y_ea;
vector[N_ctq] y_pa;
vector[N_ctq] y_sa;
vector[N_ctq] y_en;
vector[N_ctq] y_pn;


}

parameters{

vector[N_state] z_raw;
real <lower=0> lambda_0;
real gamma_abuse;
real gamma_neglect;
real <lower=0> sigma_z;
real <lower=0> sigma_pair;

vector[N_pair] u_pair;

real beta_scz;

real a_hcc;
real <lower=0> b_hcc;
real <lower=0> sigma_hcc;

real a_panss;
real b_panss;
real <lower=0> sigma_panss;

real a_pss;
real b_pss;
real <lower=0> sigma_pss;

real <lower=0> sigma_ea;
real <lower=0> sigma_pa;
real <lower=0> sigma_sa;
real <lower=0> sigma_en;
real <lower=0> sigma_pn;


real alpha_ea;
real alpha_pa;
real alpha_sa;
real alpha_en;
real alpha_pn;

matrix[2, N_subj] c_raw;
vector <lower=0> [2] sigma_c;
cholesky_factor_corr[2] L_c;

}

transformed parameters {

// Covariance matrix for CTQ latent factor c1 (abuse) and c2 (neglect)
matrix[2, N_subj] c;
c = diag_pre_multiply(sigma_c, L_c) * c_raw;

// latent subject-level factors c[1, ] and c[2, ]. Implicit Bayesian imputation. 
vector[N_subj] lambda_subj;

for (i in 1:N_subj) {
  lambda_subj[i] = lambda_0 + exp(gamma_abuse * c[1, i] + gamma_neglect * c[2, i]);
}

// dynamic latent process 
vector[N_state] z;
for (k in 1:N_state) {
    // first point in time aligned between patients and controls
    if (is_first[k] == 1) {
        z[k] = u_pair[pair_id[k]] + beta_scz * scz[k] + 1 * z_raw[k];
    // Ornstein-Uhlenbeck process (exponential return with stochastic volatility)
    } else {
        real dt = time[k] - time[k-1];
        real phi = exp(-lambda_subj[subj_id[k]] * dt);
        real sigma_dt = sigma_z * sqrt(1 - square(phi));
        z[k] = phi * z[k-1] + sigma_dt * z_raw[k];
    }
  } 


}

model{

// priors
lambda_0 ~ normal(0, 1);
gamma_abuse ~ normal(0, 1);
gamma_neglect ~ normal(0, 1);
sigma_z ~ exponential(1);
sigma_pair ~ exponential(1);
sigma_hcc ~ exponential(1);
beta_scz ~ normal(0, 1);
a_hcc ~ normal(0, 0.5);
b_hcc ~ normal(0, 1);
a_panss ~ normal(0, 0.5);
b_panss ~ normal(0, 1);
a_pss ~ normal(0, 0.5);
b_pss ~ normal(0, 1);
z_raw ~ std_normal();

u_pair ~ normal(0, sigma_pair);

sigma_ea ~ exponential(1);
sigma_pa ~ exponential(1);
sigma_sa ~ exponential(1);
sigma_en ~ exponential(1);
sigma_pn ~ exponential(1);

alpha_ea ~ normal(0, 1);
alpha_pa ~ normal(0, 1);
alpha_sa ~ normal(0, 1);
alpha_en ~ normal(0, 1);
alpha_pn ~ normal(0, 1);

to_vector(c_raw) ~ std_normal();

L_c ~ lkj_corr_cholesky(2);


// hcc emissions
for (m in 1:N_hcc) {
    y_hcc[m] ~ student_t(6, a_hcc + b_hcc * z[state_hcc[m]], sigma_hcc);
    }

// panss emissions
for (m in 1:N_panss) {
    y_panss[m] ~ normal(a_panss + b_panss * z[state_panss[m]], sigma_panss);
    }

// pss emissions
for (m in 1:N_pss) {
    y_pss[m] ~ normal(a_pss + b_pss * z[state_pss[m]], sigma_pss);
    }

// CTQ emissions 
y_ea ~ normal(alpha_ea + c[1, subj_ctq]', sigma_ea);
y_sa ~ normal(alpha_sa + c[1, subj_ctq]', sigma_sa);
y_pa ~ normal(alpha_pa + c[1, subj_ctq]', sigma_pa);
y_en ~ normal(alpha_en + c[2, subj_ctq]', sigma_en);
y_pn ~ normal(alpha_pn + c[2, subj_ctq]', sigma_pn);


}

generated quantities{


matrix[2, 2] corr = multiply_lower_tri_self_transpose(L_c); // Correlation matrix
matrix[2, 2] cov = quad_form_diag(corr, sigma_c); // (S * R * S) Covariance matrix

}

```


```{r}
#| include: false

state_df <- df_hcc %>%
  # filter(!is.na(date_total)) %>% 
  # filter(group == "HC") %>%
  # filter(kod %in% c("H-VX2J9", "H-L32SM", "H-MNQKU")) %>%
  filter_out(kod == "H-H79F8") %>% 
  group_by(kod) %>% 
  arrange(date_total, .by_group = TRUE) %>% 
  mutate(subj_id = cur_group_id(),
         first = if_else(row_number() == 1, 1, 0)) %>% 
  # select(kod:visit, panss, segment:date_total, subj_id, first, HCC0, visit_spec) %>% 
  ungroup() %>% 
  mutate(state_id = 1:n(),
         scz = if_else(group == "SCZ", 1, 0))
 
state_df_long <- state_df %>% 
  pivot_longer(c(z_log_hcc, z_panss, z_pss))

hcc_df <- state_df %>%
  filter(!is.na(HCC)) %>% 
  distinct(HCC, HCC0, state_id) %>% 
  mutate(obs_id = 1:n()) %>%
  mutate(log_hcc = log(HCC),
         z_log_hcc = scale(log_hcc)[, 1],
         avg = mean(log_hcc),
         std = sd(log_hcc))

panss_df <- state_df %>% 
  filter(!is.na(panss)) %>% 
  filter(segment == 1) %>%
  distinct(panss, state_id) %>% 
  mutate(obs_id = 1:n()) %>% 
  mutate(z_panss = scale(panss)[, 1],
         avg = mean(panss),
         std = sd(panss))



pss_df <- state_df %>% 
  filter(!is.na(pss)) %>% 
  filter(segment == 1) %>% 
  distinct(pss, state_id) %>% 
  mutate(obs_id = 1:n()) %>% 
  mutate(z_pss = scale(pss)[, 1],
         avg = mean(pss),
         std = sd(pss))

ctq_df <- state_df %>% 
  filter(!is.na(ctq_1)) %>% 
  filter(segment == 1) %>% 
  distinct(pick(contains("ctq"), state_id, subj_id)) %>% 
  mutate(obs_id = 1:n()) %>% 
  mutate(across(ctq_1:ctq_5, ~scale(.)[, 1]))


data = list(N_state = max(nrow(state_df)),
            subj_id = state_df$subj_id,
            N_pair = max(state_df$matched_pair),
            pair_id = state_df$matched_pair,
            time = state_df$date_total,
            is_first = state_df$first,
            scz = state_df$scz,
            N_hcc = max(nrow(hcc_df)),
            state_hcc = hcc_df$state_id,
            y_hcc = hcc_df$z_log_hcc,
            N_panss = max(nrow(panss_df)),
            state_panss = panss_df$state_id,
            y_panss = panss_df$z_panss,
            N_pss = max(nrow(pss_df)),
            state_pss = pss_df$state_id,
            y_pss = pss_df$z_pss,
            y_ea = ctq_df$ctq_1,
            y_pa = ctq_df$ctq_2,
            y_sa = ctq_df$ctq_3,
            y_en = ctq_df$ctq_4,
            y_pn = ctq_df$ctq_5,
            N_ctq = max(nrow(ctq_df)),
            state_ctq = ctq_df$state_id,
            subj_ctq = ctq_df$subj_id,
            N_subj = max(ctq_df$subj_id))

```



```{r}
#| output: false

library(arrow)

post <- scz_model$sample(
  data = data,
  chains = 4,
  iter_sampling = 1000,
  parallel_chains = 4,
  refresh = 0
)

pars <- post$draws(format = "df")


# # write_feather(pars, "posterior_draws.feather")
# # # 
# pars <- read_feather("posterior_draws.feather") %>%
#   as_draws_df() # Restores posterior package attributes (chains, iterations)

pred_df <- pars %>% 
  select(starts_with("z[")) %>% 
  mutate(draw = 1:nrow(.)) %>% 
  pivot_longer(-draw) %>% 
  mutate(state_id = as.numeric(str_extract(name, "\\d+"))) %>% 
  left_join(state_df, by = "state_id") %>% 
  filter(group == "SCZ") %>% 
  mutate(time_bin = round(date_total)) %>%
  # filter(time_bin %in% c(-10)) %>%
  arrange(draw, kod, state_id)

pred_df_hc <- pars %>% 
  select(starts_with("z[")) %>% 
  mutate(draw = 1:nrow(.)) %>% 
  pivot_longer(-draw) %>% 
  mutate(state_id = as.numeric(str_extract(name, "\\d+"))) %>% 
  left_join(state_df, by = "state_id") %>% 
  mutate(time_bin = round(date_total)) %>%
  # filter(time_bin %in% c(-10)) %>%
  arrange(draw, kod, state_id)



ctq_draws <- pars %>%
  # Adjust the string regex if your variables are named c_raw or c
  select(starts_with("c[")) %>% 
  mutate(draw = 1:nrow(.)) %>%
  pivot_longer(-draw, names_to = "c_name", values_to = "ctq_factor") %>%
  # Parse out the factor dimension (1 = Abuse, 2 = Neglect) and the subject index
  mutate(
    dimension = as.numeric(str_match(c_name, "c\\[(\\d+),\\s*(\\d+)\\]")[, 2]),
    subj_id   = as.numeric(str_match(c_name, "c\\[(\\d+),\\s*(\\d+)\\]")[, 3])
  ) %>%
  # Let's focus on c[1] (Abuse) for this stratification. Change to 2 for neglect.
  # filter(dimension == 1) %>% 
  select(draw, subj_id, ctq_factor) %>% 
  # slice_sample(n = 20000)
  identity()

# --- Step 2: Merge everything into the primary dataframe ---
pred_df_stratified <- pars %>% 
  select(starts_with("z[")) %>% 
  mutate(draw = 1:nrow(.)) %>%
  filter(draw %in% ctq_draws$draw) %>% 
  pivot_longer(starts_with("z["), names_to = "z_name", values_to = "z_value") %>% 
  mutate(state_id = as.numeric(str_extract(z_name, "\\d+"))) %>% 
  # Join state details to find out which subject (subj_id) each state belongs to
  left_join(state_df, by = "state_id") %>% 
  # Join the draw-specific latent CTQ score for that specific subject
  left_join(ctq_draws, by = c("draw", "subj_id"))

# --- Step 3: Define the cut categories ---
pred_df_data <- pred_df_stratified %>% 
  mutate(
    time_bin = round(date_total),
    # Bin the continuous posterior draws of CTQ into Low/Medium/High groups
    ctq_tier = cut(
      ctq_factor, 
      breaks = c(-Inf, -0.75, 0.75, Inf), 
      labels = c("Low Trauma", "Moderate Trauma", "High Trauma")
    )
  ) %>% 
  filter(!is.na(ctq_tier)) %>% # Drop subjects without valid data/mappings if any
  arrange(draw, kod, state_id)

state_df_long <- state_df %>% 
  pivot_longer(c(z_log_hcc, z_panss, z_pss))

pred_diff <- pred_df_hc %>%
  # filter(group == "SCZ")
  # filter(draw %in% c(1, 2),
  #        kod == c("H-2A3N2", "H-42MDQ")) %>%
  # filter(visit == "V1") %>%
  select(draw, value, time_bin, group) %>%
  group_by(group, time_bin, draw) %>%
  mutate(pid = 1:n()) %>%
  ungroup() %>% 
  pivot_wider(names_from = group, values_from = value) %>% 
  mutate(diff = HC - SCZ) %>%
  group_by(time_bin) %>% 
  summarise(
    mean = mean(diff, na.rm = TRUE),
    lower = quantile(diff, 0.1, na.rm = TRUE),
    upper = quantile(diff, 0.9, na.rm = TRUE),
    sign = mean > 0,
    p_above = mean(diff > 0, na.rm = TRUE)
  ) %>%
  mutate(
    upper_pos = pmax(upper, 0),
    lower_pos = pmax(lower, 0),
    upper_neg = pmin(upper, 0),
    lower_neg = pmin(lower, 0)
  ) %>% 
  ungroup() %>%
  mutate(avg_p_above = mean(p_above, na.rm = TRUE))

```

#### Model diagnostics and checks

The model was rigorously evaluated for convergence using four Markov chains with 4000 iterations each. The Rhat diagnostic was used to assess convergence, with a value below 1.01 suggesting adequate convergence. 
The posterior predictive checks follow.  


```{r}
#| include: false

library(tidyverse)
library(tidybayes)
library(posterior)


# Number of posterior draws used for PPCs.
# 500 is usually plenty for figures and keeps the objects manageable.
n_ppc <- 150

# Draws needed for all PPCs
draws_all <- pars

# Randomly retain posterior draws
draw_ids <- sample(
  unique(draws_all$.draw),
  size = n_ppc)

draws <- draws_all |>
  filter(.draw %in% draw_ids)
```

```{r}
#| include: false

state_info <- tibble(
  state = seq_len(data$N_state),
  subj = data$subj_id,
  pair = data$pair_id,
  time = data$time,
  first = data$is_first,
  scz = data$scz
)

hcc_obs <- tibble(
  obs = seq_len(data$N_hcc),
  state = data$state_hcc,
  y = data$y_hcc
) |>
  left_join(state_info, by = "state")

panss_obs <- tibble(
  obs = seq_len(data$N_panss),
  state = data$state_panss,
  y = data$y_panss
) |>
  left_join(state_info, by = "state")

pss_obs <- tibble(
  obs = seq_len(data$N_pss),
  state = data$state_pss,
  y = data$y_pss
) |>
  left_join(state_info, by = "state")

ctq_obs <- tibble(
  obs = seq_len(data$N_ctq),
  subj = data$subj_ctq,
  EA = data$y_ea,
  PA = data$y_pa,
  SA = data$y_sa,
  EN = data$y_en,
  PN = data$y_pn
)

z_draws <- draws |>
  select(.draw, starts_with("z[")) |>
  pivot_longer(
    - .draw,
    names_to = "state",
    values_to = "z"
  ) |>
  mutate(
    state = readr::parse_number(state)
  ) |>
  left_join(state_info, by = "state")

c_draws <- draws |>
  select(.draw, starts_with("c[")) |>
  pivot_longer(
    - .draw,
    names_to = "parameter",
    values_to = "c"
  ) |>
  extract(
    parameter,
    into = c("factor", "subj"),
    regex = "c\\[(\\d+),(\\d+)\\]",
    convert = TRUE
  ) |>
  mutate(
    factor = case_when(
      factor == 1 ~ "abuse",
      factor == 2 ~ "neglect"
    )
  )

lambda_draws <- draws |>
  select(.draw, starts_with("lambda_subj[")) |>
  pivot_longer(
    - .draw,
    names_to = "subj",
    values_to = "lambda"
  ) |>
  mutate(
    subj = readr::parse_number(subj)
  )

hcc_par <- draws |>
  select(
    .draw,
    a_hcc,
    b_hcc,
    sigma_hcc
  )

hcc_rep <- z_draws |>
  select(.draw, state, subj, pair, time, scz, z) |>
  inner_join(
    hcc_obs |>
      select(obs, state),
    by = "state"
  ) |>
  left_join(hcc_par, by = ".draw") |>
  mutate(
    y_rep = a_hcc +
      b_hcc * z +
      sigma_hcc * rt(n(), df = 6)
  )

panss_par <- draws |>
  select(
    .draw,
    a_panss,
    b_panss,
    sigma_panss
  )

panss_rep <- z_draws |>
  select(.draw, state, subj, pair, time, scz, z) |>
  inner_join(
    panss_obs |>
      select(obs, state),
    by = "state"
  ) |>
  left_join(panss_par, by = ".draw") |>
  mutate(
    y_rep = rnorm(
      n(),
      mean = a_panss + b_panss * z,
      sd = sigma_panss
    )
  )

pss_par <- draws |>
  select(
    .draw,
    a_pss,
    b_pss,
    sigma_pss
  )

pss_rep <- z_draws |>
  select(.draw, state, subj, pair, time, scz, z) |>
  inner_join(
    pss_obs |>
      select(obs, state),
    by = "state"
  ) |>
  left_join(pss_par, by = ".draw") |>
  mutate(
    y_rep = rnorm(
      n(),
      mean = a_pss + b_pss * z,
      sd = sigma_pss
    )
  )

ctq_par <- draws |>
  select(
    .draw,
    alpha_ea, alpha_pa, alpha_sa,
    alpha_en, alpha_pn,
    sigma_ea, sigma_pa, sigma_sa,
    sigma_en, sigma_pn
  )

ctq_latent <- c_draws |>
  select(.draw, factor, subj, c) |>
  pivot_wider(
    names_from = factor,
    values_from = c
  )

ctq_rep <- ctq_obs |>
  select(obs, subj) |>
  crossing(
    tibble(.draw = draw_ids)
  ) |>
  left_join(ctq_latent, by = c(".draw", "subj")) |>
  left_join(ctq_par, by = ".draw") |>
  mutate(
    EA_rep = rnorm(
      n(), alpha_ea + abuse, sigma_ea
    ),
    PA_rep = rnorm(
      n(), alpha_pa + abuse, sigma_pa
    ),
    SA_rep = rnorm(
      n(), alpha_sa + abuse, sigma_sa
    ),
    EN_rep = rnorm(
      n(), alpha_en + neglect, sigma_en
    ),
    PN_rep = rnorm(
      n(), alpha_pn + neglect, sigma_pn
    )
  )


```

```{r}
#| include: false

plot_ppc_density <- function(rep_data, observed, variable_rep,
                             variable_obs, xlab) {

  obs <- observed |>
    transmute(
      value = .data[[variable_obs]]
    )

  rep <- rep_data |>
    transmute(
      .draw,
      value = .data[[variable_rep]]
    )

  ggplot() +
    geom_density(
      data = rep,
      aes(x = value, group = .draw),
      alpha = 0.04
    ) +
    geom_density(
      data = obs,
      aes(x = value),
      linewidth = 1,
      color = "red"
    ) +
    labs(
      x = xlab,
      y = "Density"
    ) 
}
```

#### Posterior predictive checks

##### Density

```{r}
#| label: fig-ppc1
#| fig-cap: "Posterior predictive check of HCC. The red line shows the observed values."

plot_ppc_density(
  hcc_rep,
  hcc_obs,
  "y_rep",
  "y",
  "Standardized log HCC"
)
```

```{r}
#| label: fig-ppc2
#| fig-cap: "Posterior predictive check of PANSS. The red line shows the observed values."

plot_ppc_density(
  panss_rep,
  panss_obs,
  "y_rep",
  "y",
  "Standardized PANSS"
)
```

```{r}
#| label: fig-ppc3
#| fig-cap: "Posterior predictive check of PSS. The red line shows the observed values."

plot_ppc_density(
  pss_rep,
  pss_obs,
  "y_rep",
  "y",
  "Standardized PSS"
)
```

```{r}
ctq_rep_long <- ctq_rep |>
  select(.draw, subj, ends_with("_rep")) |>
  pivot_longer(
    ends_with("_rep"),
    names_to = "scale",
    values_to = "y_rep"
  ) |>
  mutate(
    scale = str_remove(scale, "_rep$")
  )

ctq_obs_long <- ctq_obs |>
  pivot_longer(
    c(EA, PA, SA, EN, PN),
    names_to = "scale",
    values_to = "y"
  )
```


```{r}
#| label: fig-ppc4
#| fig-cap: "Posterior predictive check of the latent CTQ. The red line shows the observed values."

ggplot(ctq_rep_long, aes(x = y_rep, group = .draw)) +
  geom_density(alpha = 0.04) +
  geom_density(
    data = ctq_obs_long,
    aes(x = y),
    inherit.aes = FALSE,
    linewidth = 1,
    color = "red"
  ) +
  facet_wrap(~ scale, scales = "free") +
  labs(
    x = "CTQ score",
    y = "Density"
  ) 

```

##### Statistics

```{r}
ppc_statistics <- function(rep_data, observed) {

  rep_stats <- rep_data |>
    group_by(.draw) |>
    summarise(
      mean = mean(y_rep),
      sd = sd(y_rep),
      q05 = quantile(y_rep, .05),
      median = median(y_rep),
      q95 = quantile(y_rep, .95),
      min = min(y_rep),
      max = max(y_rep),
      .groups = "drop"
    )

  obs_stats <- observed |>
    summarise(
      mean = mean(y),
      sd = sd(y),
      q05 = quantile(y, .05),
      median = median(y),
      q95 = quantile(y, .95),
      min = min(y),
      max = max(y)
    ) |>
    pivot_longer(
      everything(),
      names_to = "statistic",
      values_to = "observed"
    )

  rep_stats |>
    pivot_longer(
      - .draw,
      names_to = "statistic",
      values_to = "replicated"
    ) |>
    left_join(obs_stats, by = "statistic")
}

```

```{r}
#| label: fig-ppc5
#| fig-cap: "Posterior predictive check of selected statistics of HCC. The red line shows the observed values."

hcc_stats <- ppc_statistics(hcc_rep, hcc_obs)

ggplot(hcc_stats, aes(x = replicated)) +
  geom_density() +
  geom_vline(
    aes(xintercept = observed),
    linewidth = 1,
    color = "red"
  ) +
  facet_wrap(~ statistic, scales = "free") +
  labs(
    x = NULL,
    y = "Posterior predictive density"
  )

```


```{r}
#| label: fig-ppc6
#| fig-cap: "Posterior predictive check of selected statistics of PANSS. The red line shows the observed values."

panss_stats <- ppc_statistics(panss_rep, panss_obs)

ggplot(panss_stats, aes(x = replicated)) +
  geom_density() +
  geom_vline(
    aes(xintercept = observed),
    linewidth = 1,
        color = "red"
  ) +
  facet_wrap(~ statistic, scales = "free") +
  labs(
    x = NULL,
    y = "Posterior predictive density"
  )
```


```{r}
#| label: fig-ppc7
#| fig-cap: "Posterior predictive check of selected statistics of PSS. The red line shows the observed values."

pss_stats <- ppc_statistics(pss_rep, pss_obs)

ggplot(pss_stats, aes(x = replicated)) +
  geom_density() +
  geom_vline(
    aes(xintercept = observed),
    linewidth = 1,    
    color = "red"
  ) +
  facet_wrap(~ statistic, scales = "free") +
  labs(
    x = NULL,
    y = "Posterior predictive density"
  )

```













